MSc Statistical overview

### 2018-01-29 MScFigures.Rmd
Added simplified analysis and data visualizations -- Rob

```{r Load_packages and functions, message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(MASS)
library(MuMIn)
library(emmeans)
library(boot)
library(brms)
library(loo)
library(fitdistrplus)
library(mefa)
library(ggbiplot)

# theme_simple() a homemade function from the Colautti lab (2017) to standardize figures
theme_simple <- function (base_size = 16, base_family = "") {
  theme_classic(base_size = base_size, base_family = base_family) %+replace%
    theme(
      axis.text = element_text(colour = "black"),
      axis.title.x = element_text(size=18),
      axis.text.x = element_text(size=16, face= "italic"),
      axis.title.y = element_text(size=18, angle=90), 
      axis.text.y = element_text(size=16),
      axis.ticks = element_blank(),
      panel.background = element_rect(fill="white"),
      panel.border = element_blank(),
      plot.title=element_text(face="bold", size=24)
      #      legend.position="none"
    )
}

overdisp_fun <- function(model) {
  ## number of variance parameters in an n-by-n variance-covariance matrix
  vpars <- function(m) {
    nrow(m) * (nrow(m) + 1)/2
  }
  # The next two lines calculate the residual degrees of freedom
  model.df <- sum(sapply(VarCorr(model), vpars)) + length(fixef(model))
  rdf <- nrow(model.frame(model)) - model.df
  # extracts the Pearson residuals
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  prat <- Pearson.chisq/rdf
  # Generates a p-value. If less than 0.05, the data are overdispersed.
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = prat, rdf = rdf, p = pval)
}
```

@@

### Load/ Modify data: Co-occurence
```{r Introduce co-occurence dataset, message=FALSE, warning=FALSE}
mdata <- read.csv("Assessing_Oct_20.csv",stringsAsFactors=F)
indiv <- mdata$CODE[!mdata$CODE==""]
tdata <- mdata[grep("[0-9]+",mdata$Cross),]
nrow(tdata)/length(indiv)
# This is not the same because some individuals just had the totals not the cooccurence data
indiv2 <- tdata$CODE[!tdata$CODE==""]
nrow(tdata)/length(indiv2)
indiv2 <- data.frame(indiv2)
indiv2 <- mefa:::rep.data.frame(indiv2,each=100)
fdata <- cbind(indiv2,tdata)
fdata <- fdata[,-c(2,3,4,9,14,17,18)]

# Cleaning up the occurence data
fdata[is.na(fdata)] <- 0
fdata$Lesion<-rowSums(fdata[,2:4])
fdata[fdata$Lesion>1,]
fdata$Mycorrhiza<-rowSums(fdata[,6:8])
fdata[fdata$Mycorrhiza>1,]
fdata <- fdata[,-c(2:4,6:8)]

@@

# Reviewing the hypothesis

## Histology data Hypothesis
### Hypothesis 1

Mycorrhizal colonization protects roots from lesions and herbivory.
```{r echo=FALSE}
y<-data.frame(c(rnorm(50, mean=5, sd=8),rnorm(50, mean=50, sd=8)))
colnames(y)<-c("y")
y$y<-sort(y$y, decreasing = T)
x<-data.frame(c(rnorm(50, mean=5, sd=8),rnorm(50, mean=50, sd=8)))
colnames(x)<-c("x")
x$x<-sort(x$x, decreasing = F)
df <- cbind(x,y)
colnames(df)<-c("x","y")

ggplot(df, aes(x,y))+geom_smooth(method="lm", se=F)+geom_point()+
  theme_simple()+labs(x="Mycorrhizal Colonization (%)", y="Lesions (%)")+
  theme(axis.text.x = element_blank(),axis.text.y = element_blank())
```

@@

A. petiolata invasion reduces mycorrhizal colonization and increases pathogen colonization and herbivory
```{r echo=FALSE}
InOut<- data.frame(c(rep("I", times=50),rep("O", times=50)))
colnames(InOut)<-c("Location")
df <- cbind(df,InOut)

ggplot(df, aes(x,y, colour=Location))+geom_smooth(method="lm", se=F)+geom_jitter()+
  theme_simple()+labs(x="Mycorrhizal Colonization (%)", y="Lesions (%)")+
  theme(axis.text.x = element_blank(),axis.text.y = element_blank())
```
### Hypothesis 3

Efficiency of mycorrhizal protection is reduced in the presence of A. petiolata
```{r echo=FALSE}
y<-data.frame(c(rnorm(50, mean=5, sd=8),rep(40, times=10),rep(41, times=10),rep(42, times=10),rep(40.5, times=10),rep(39, times=10)))
colnames(y)<-c("y")
y$y<-sort(y$y, decreasing = T)
x<-data.frame(c(rnorm(50, mean=5, sd=8),rnorm(50, mean=50, sd=8)))
colnames(x)<-c("x")
x$x<-sort(x$x, decreasing = F)
df <- cbind(x,y)
colnames(df)<-c("x","y")

InOut<- data.frame(c(rep("I", times=50),rep("O", times=50)))
colnames(InOut)<-c("Location")
df <- cbind(df,InOut)

ggplot(df, aes(x,y, colour=Location))+geom_smooth(method="lm", se=F)+geom_jitter()+
  theme_simple()+labs(x="Mycorrhizal Colonization (%)", y="Lesions (%)")+
  theme(axis.text.x = element_blank(),axis.text.y = element_blank())
```

@@

@@

@@

@@

```{r Co-occurence model using corr}
# calculating the r
indivCODE <- unique(fdata$indiv2)
dfcorr <-data.frame(indivCODE=c(0),r=c(0))
for (i in 1:length(indivCODE)){
  dat1<-NULL
  dat1<-filter(fdata, fdata$indiv2==indivCODE[i])
  dfcorr[i,1] <- indivCODE[i]
  dfcorr[i,2] <- cor(as.numeric(dat1$Mycorrhiza),as.numeric(dat1$Lesion))
}
dfcorr %>% drop_na()->dfcorr

#Normal?
car::qqp(dfcorr$r, "norm")
shapiro.test(dfcorr$r)
hist(dfcorr$r)
# Cannot be normal because it is only between -1 and 1 and our mean is at 0.
descdist(dfcorr$r, discrete = FALSE, boot = 500)

>>>>

```{r Load_packages and functions, message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(MASS)
library(MuMIn)
library(emmeans)
library(boot)
library(brms)
library(loo)
library(fitdistrplus)
library(mefa)
library(ggbiplot)
```

## KAT, PLEASE BREAK UP YOUR CODE INTO SMALL CHUNCKS AND EXPLAIN WHAT EACH IS DOING. 
Load a custome theme for clean, readable graphs
```{r echo=F}
# theme_simple() a homemade function from the Colautti lab (2017) to standardize figures
theme_simple <- function (base_size = 16, base_family = "") {
  theme_classic(base_size = base_size, base_family = base_family) %+replace%
    theme(
      axis.text = element_text(colour = "black"),
      axis.title.x = element_text(size=18),
      axis.text.x = element_text(size=16, face= "italic"),
      axis.title.y = element_text(size=18, angle=90), 
      axis.text.y = element_text(size=16),
      axis.ticks = element_blank(),
      panel.background = element_rect(fill="white"),
      panel.border = element_blank(),
      plot.title=element_text(face="bold", size=24)
      #      legend.position="none"
    )
}
```

## EXPLAIN WHAT THIS OVERDISPERSION FUNCTION IS FOR...
```{r echo=F}
overdisp_fun <- function(model) {
  ## number of variance parameters in an n-by-n variance-covariance matrix
  vpars <- function(m) {
    nrow(m) * (nrow(m) + 1)/2
  }
  # The next two lines calculate the residual degrees of freedom
  model.df <- sum(sapply(VarCorr(model), vpars)) + length(fixef(model))
  rdf <- nrow(model.frame(model)) - model.df
  # extracts the Pearson residuals
  rp <- residuals(model, type = "pearson")
  Pearson.chisq <- sum(rp^2)
  prat <- Pearson.chisq/rdf
  # Generates a p-value. If less than 0.05, the data are overdispersed.
  pval <- pchisq(Pearson.chisq, df = rdf, lower.tail = FALSE)
  c(chisq = Pearson.chisq, ratio = prat, rdf = rdf, p = pval)
}
```

@@

### Load/ Modify data: Co-occurence
```{r Introduce co-occurence dataset, message=FALSE, warning=FALSE}
mdata <- read.csv("Assessing_Oct_20.csv",stringsAsFactors=F)
indiv <- mdata$CODE[!mdata$CODE==""]
tdata <- mdata[grep("[0-9]+",mdata$Cross),]
nrow(tdata)/length(indiv)

# This is not the same because some individuals just had the totals not the cooccurence data
## THE COMMENT IS NOT CLEAR, EXPLAIN...
indiv2 <- tdata$CODE[!tdata$CODE==""]
nrow(tdata)/length(indiv2)
indiv2 <- data.frame(indiv2)
indiv2 <- mefa:::rep.data.frame(indiv2,each=100)
fdata <- cbind(indiv2,tdata)
fdata <- fdata[,-c(2,3,4,9,14,17,18)]

# Cleaning up the occurence data
## CLEANING UP HOW?
fdata[is.na(fdata)] <- 0
fdata$Lesion<-rowSums(fdata[,2:4])
fdata[fdata$Lesion>1,]
fdata$Mycorrhiza<-rowSums(fdata[,6:8])
fdata[fdata$Mycorrhiza>1,]
fdata <- fdata[,-c(2:4,6:8)]

@@

# Reviewing of hypotheses:

## Question 1: Is there evidence that Mycorrhizal colonization provides protection against root pathogen damage?

## Question 2: Do these effects vary among species?

### VERY NICE PREDICTION FIGURES! PERSONALLY, I WOULD MAKE THE DOTS BIGGER AND BIVARIATE NORMAL (SEE CHANGES):

Mycorrhizal colonization protects roots from lesions and herbivory.
```{r echo=FALSE}
set.seed(3242)
y<-data.frame(rnorm(100,sd=7))
colnames(y)<-c("y")
y$y<-sort(y$y)
x<-data.frame(-y+rnorm(100))
colnames(x)<-c("x")
df <- cbind(x,y)
colnames(df)<-c("x","y")
df$Location <- c(rep("Out",nrow(df)/2),rep("In",nrow(df)/2))

ggplot(data=df, aes(x,y))+geom_smooth(method="lm", se=F)+geom_point(size=4,alpha=0.3)+
  theme_simple()+labs(x="Mycorrhizal Colonization (%)", y="Lesions (%)")+
  theme(axis.text.x = element_blank(),axis.text.y = element_blank())
```

@@

A. petiolata invasion reduces mycorrhizal colonization and increases pathogen colonization and herbivory
```{r echo=FALSE}
ggplot(df, aes(x,y, colour=Location))+geom_smooth(method="lm", se=F)+
  geom_point(size=4,alpha=0.3)+theme_simple()+
  labs(x="Mycorrhizal Colonization (%)", y="Lesions (%)")+
  theme(axis.text.x = element_blank(),axis.text.y = element_blank())
```
### Hypothesis 3

Cycorrhizal colonization is reduced in the presence of A. petiolata
```{r echo=FALSE}
df$y2<-df$y
df$y2[df$Location=="In"]<-df$y2[df$Location=="In"]+10
ggplot(df, aes(x,y2, colour=Location))+geom_smooth(method="lm", se=F)+
  geom_point(size=4,alpha=0.3)+
  theme_simple()+labs(x="Mycorrhizal Colonization (%)", y="Lesions (%)")+
  theme(axis.text.x = element_blank(),axis.text.y = element_blank())
```

### Hypothesis 4

Mycorrhizal protection is completely lost in the presence of A. petiolata
```{r echo=FALSE}
df$y2<-df$y
set.seed (238.4)
df$y2[df$Location=="In"]<-rnorm(nrow(df)/2,mean=10)
ggplot(df, aes(x,y2, colour=Location))+geom_smooth(method="lm", se=F)+
  geom_point(size=4,alpha=0.3)+
  theme_simple()+labs(x="Mycorrhizal Colonization (%)", y="Lesions (%)")+
  theme(axis.text.x = element_blank(),axis.text.y = element_blank())
```

@@

## IMPORTANT: BEFORE RUNNING COMPLICATED MODELS: (1) TAKE A LOOK AT THE DATA, (2) RUN SIMPLE MODELS -- SEE IF IT IS EVEN WORTH YOUR TIME & EFFORT

@@

### Data exploration:

Calculate plant-level Mycchorizae, pathogens and herbivory:

```{r}
# Fix column characteristics for plotting and analysis
fdata$indiv2<-as.factor(fdata$indiv2)
fdata$Mycorrhiza<-as.numeric(paste(fdata$Mycorrhiza))
fdata$Lesion<-as.numeric(paste(fdata$Lesion))
fdata$Herbivory<-as.numeric(paste(fdata$Herbivory))
pct<-function(x){
  sum(x)/100
}

# Calculate plant-level averages
fdat_ind<-aggregate(fdata[,c("Mycorrhiza","Lesion","Herbivory")],
                    by=list(indiv2=fdata$indiv2,pop=fdata$pop,loc=fdata$loc,species=fdata$species),
                    pct)

## SIMPLE LINEAR MODEL AND VISUALIZATION:

### Mycorrhia model results

  1. Mycorrhiza colonization is significantly higher outside patches of A. petiolata than inside
  2a. Species vary in Mycorrhiza colonization, with highest colonization in SC and others relatively similar. 
  2b. Effect of A. petiolata on Mycorrhiza colonization differs among species, with more pronounced effects for GR and MR, and will little effect for GA
  3. Effect of A. petiolata on Mycorrhiza colonization varies by population, but not as strongly as among species

```{r}
## Mycorrhiza model (NOTE: No loc*species because not all species present in all populations)
base.Myc<-lmer(Mycorrhiza~species+loc+species:loc+(1|pop)+(1|pop:loc),data=fdat_ind)
noint.Myc<-lmer(Mycorrhiza~species+loc+(1|pop)+(1|pop:loc),data=fdat_ind)
anova(base.Myc,noint.Myc)
base.Myc ## Inspect coefficients

ggplot(fdat_ind,aes(species,Mycorrhiza,colour=loc))+geom_boxplot()+theme_simple()
ggplot(fdat_ind,aes(pop,Mycorrhiza,colour=loc))+geom_boxplot()+theme_simple()
```

### Lesions model results

  1. Root lesions increase with mycorrhiza colonization, contrary to predictions
  2. Effect of Mycorrhiza on Lesions depends strongly on species (Myc*species interactions)

```{r}
## Mycorrhiza model (NOTE: No loc*species because not all species present in all populations)
base.Les<-lmer(Lesion~Mycorrhiza*species*loc+(1|pop/loc),data=fdat_ind)
norint.Les<-lmer(Lesion~Mycorrhiza*species*loc+(1|pop),data=fdat_ind)
anova(base.Les,norint.Les)
norint.Les # inspect coefficients

Les2<-lmer(Lesion~Mycorrhiza*species*loc-Mycorrhiza:species:loc+(1|pop),data=fdat_ind)
anova(norint.Les,Les2)

Les3a<-lmer(Lesion~Mycorrhiza*species+Mycorrhiza*loc+(1|pop),data=fdat_ind)
Les3b<-lmer(Lesion~Mycorrhiza*species+species*loc+(1|pop),data=fdat_ind)
Les3c<-lmer(Lesion~species*loc+Mycorrhiza*loc+(1|pop),data=fdat_ind)
anova(norint.Les,Les3a) # TEST species*loc 
anova(norint.Les,Les3b) # TEST Myc*loc 
anova(norint.Les,Les3c) # TEST Myc*species

ggplot(fdat_ind,aes(Mycorrhiza,Lesion))+facet_wrap(~species)+
  geom_point(aes(col=loc))+theme_simple()+geom_smooth(method="lm")
```

@@

### GIVEN THE NEW ANALYSIS ABOVE, THIS ISN'T WORTHWILE BECAUSE THERE IS NO SIGNIFICANT Myc:species:loc

```{r}
# calculating the r
indivCODE <- unique(fdata$indiv2)
dfcorr <-data.frame(indivCODE=c(0),r=c(0))
for (i in 1:length(indivCODE)){
  dat1<-NULL
  dat1<-filter(fdata, fdata$indiv2==indivCODE[i])
  dfcorr[i,1] <- indivCODE[i]
  dfcorr[i,2] <- cor(as.numeric(dat1$Mycorrhiza),as.numeric(dat1$Lesion))
}
dfcorr %>% drop_na()->dfcorr
```

@@

```{r Co-occurence model using corr}
#Normal?
car::qqp(dfcorr$r, "norm")
shapiro.test(dfcorr$r)
hist(dfcorr$r)
# Cannot be normal because it is only between -1 and 1 and our mean is at 0.
descdist(dfcorr$r, discrete = FALSE, boot = 500)
